Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-05-03 concurrency/CONCUR 2023/paper.tex
more spacing fixes

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
			\right\rangle
		}\\
		\texttt{traverse(pn, k)} \\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}

@@

We next consider the give-up template, which uses an \emph{optimistic concurrency control} approach, acquiring fewer locks at the cost of sometimes having to recover from synchronization errors. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time. This means that a conflicting operation may invalidate our traversal, for instance by moving the next node to another part of the data structure before we acquire its lock. To guard against this, the \lstinline{traverse} function must explicitly check whether the target key is in the range of the current node to ensure that it is still on the right path. If a check fails, we give up and start the traversal over from the root node. The give-up template performs well in scenarios where operations generally do not conflict, either because they are mostly to independent parts of the data structure or because they mostly do not delete or relocate nodes.

@@

Figure \ref{traverse_giveup_a} depicts the give-up template algorithm as originally presented. In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors. Logically, this range is the same as the range in the lock-coupling template, but that range was a ghost-state construct that only appeared in the proofs; in the give-up template, the range must be stored in memory and checked in the code. If \lstinline{k} is outside the node's range (e.g. because the node has been relocated), the search is restarted from the root node \lstinline{r}. As in the previous section, we implement this in C with a loop: in each iteration, we acquire the lock on the current node, check that \lstinline{k} is in range, and then use \lstinline{findNext} as above, releasing the lock before we move to the next node. If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}). The \texttt{insert} operation (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version, with the one difference that it does not need to acquire a lock before calling \texttt{traverse}.

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings the key ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to as many target data structures as possible without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t} in our C implementation:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
The specification of \lstinline{inRange} is:
	\begin{mathpar}
		% inRange for give-up
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2}  
			\end{array}
			\right\}
		}
		\\ 
		\texttt{inRange(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res.} \ 
			\begin{array}{l}
				\texttt{pn->n->min}\mapsto \mathsf{n_1} \ \ast \ \texttt{pn->n->max}\mapsto\mathsf{n_2} \ \ast  \ 
				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\mathsf{n_1} < \texttt{k} < \mathsf{n_2})\ \mathsf{else}\ (\texttt{k} \leq \mathsf{n_1} \lor \texttt{k} \geq \mathsf{n_2}))
			\end{array}
			\right\}
		}
	\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range \texttt{(min, max)} associated with the node \texttt{n}, and returns 0 or 1 accordingly.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\\ 
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast\ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \ \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

Our target data structure is a binary search tree (BST) implemented in C. A lock-coupling BST was already written verified as a demonstration of VST's concurrency capabilities~\cite{bst-conc}. We refactored it into \lstinline{findNext}, \lstinline{traverse}, \lstinline{insertOp}, etc. as required by the template approach, and then re-verified it using the template; we then replaced \lstinline{traverse} with the give-up version and verified the resulting give-up BST as well. We were able to obtain verified \lstinline{insert} and \lstinline{lookup} operations using this technique; for \texttt{delete}, we found a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}. %The complete proofs in VST can be found online at \href{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}{\color{blue}{https://github.com/PrincetonUniversity/DeepSpecDB/tree/master/concurrency/templates}}; we discuss the overall development in Section \ref{proof_mech}.

@@

\begin{figure}[h]
	\centering
	%\wm{I think this should be \lstinline{pn->n->t} $\mapsto$ \lstinline{(k',v', l, r)}, but I'm not sure.}
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  \texttt{n'.}
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)}  \\\  
				\ast \ (\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{l} = \texttt{n'} \land \texttt{k} < \texttt{k'}) \lor (\texttt{r} = \texttt{n'} \land \texttt{k} > \texttt{k'})\ \mathsf{else}\ (\texttt{n} = \texttt{n'} \land \texttt{k} = \texttt{k'}))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)} %\wm{Shouldn't pn have two fields, p and n?}
	\label{fig:findNext_lock}
\end{figure}

@@

%\wm{Should we recap the definition of the \lstinline{node} struct here?} 
The contents of a BST node are defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}
The BST's \lstinline{findNext} function (Figure \ref{findnext}) indicates whether the target key is in the current node, its left subtree, or its right subtree. The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes. The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0. The proof is straightforward. %\wm{We should show the actual code for \lstinline{findNext} here!}

\begin{figure}[ht]
	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
	\caption{The \lstinline{findNext} method for binary search tree (all templates) written in C}
	\label{findnext}
\end{figure}

>>>>

%\wm{The fonts are still off here---for instance, p, n, and n' are mathematical variables, not references to the C code. Fix if we have time.}
The \lstinline{traverse} function can then be specified as follows:
%\than{Try to avoid using specific name of $\mathrm{tree\_rep}$ for public condition, I suggested to use $\mathrm{Node}$ for generic purpose.} 
	\begin{mathpar}
		% spec of inRange
		{\color{blue}
			\forall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
			\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ m \
			\right\rangle
		}\vspace{-6pt}\\
		\vspace{-6pt}\texttt{traverse(pn, k)} \\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast \ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
The local precondition of \lstinline{traverse} includes both the node handle $\nodeboxrep(n)$ and its contents $\mathsf{R}(n)$, indicating that a thread should already hold $n$'s lock before calling \lstinline{traverse}. The output of \lstinline{traverse} is a new node $n'$ in the data structure such that the key \texttt{k} falls within the range of $n'$. The local postcondition then includes the handle of the new node $\nodeboxrep(n')$, its contents $\mathsf{R}(n')$, and a Boolean variable $\mathit{res}$ indicating whether \lstinline{traverse} found an empty node or a node with key \lstinline{k}.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \  \nodeboxrep(n)  \ \ast \\  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}

@@

We next consider the give-up template, which uses an \emph{optimistic concurrency control} approach, acquiring fewer locks at the cost of sometimes having to recover from synchronization errors. Unlike the lock-coupling template, which maintains locks during traversal between nodes, the give-up template only acquires a lock just before operating on a node, and holds at most one lock at any time. This means that a conflicting operation may invalidate a traversal, for instance by moving the next node to another part of the data structure before we acquire its lock. To guard against this, the \lstinline{traverse} function must explicitly check whether the target key is in the range of the current node. If a check fails, we give up and start the traversal over from the root node. The give-up template performs well in scenarios where operations generally do not conflict, either because they are on independent parts of the data structure or because they do not delete or relocate nodes.

@@

Figure \ref{traverse_giveup_a} shows the give-up template algorithm as originally presented. In addition to \lstinline{findNext}, the \lstinline{traverse} function uses a helper function called \lstinline{inRange}, which determines whether the key value $\texttt{k}$ falls within the range of keys held in node $\texttt{n}$ and its successors. Logically, this range is the same as the range in the lock-coupling template, but that range was a ghost-state construct that only appeared in the proofs; in the give-up template, the range must be stored in memory and checked in the code. If \lstinline{k} is outside the node's range (e.g. because the node has been relocated), the search is restarted from the root node \lstinline{r}. As in the previous section, we implement this in C with a loop: in each iteration, we acquire the lock on the current node, check that \lstinline{k} is in range, and then use \lstinline{findNext} as above, releasing the lock before we move to the next node. If the \lstinline{inRange} call fails, we release our current lock and return to the root node by setting the current node to the root pointer \lstinline{p} (lines 21-22 in Figure \ref{traverse_giveup_b}). The give-up version of the \texttt{insert} operation (see Figure \ref{insert_giveup}) is almost identical to the lock-coupling version, except that it does not acquire a lock before calling \texttt{traverse}.

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {node *t; lock_t *lock; int min; int max;} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
	\begin{mathpar}
		% inRange for give-up
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2
			\end{array}
			\right\}
		}
		\vspace{-6pt}\\ \vspace{-6pt}
		\texttt{inRange(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res.} \ 
			\begin{array}{l}
				\texttt{pn->n->min}\mapsto n_1 \ \ast \ \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
				\\(\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2))
			\end{array}
			\right\}
		}
	\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

@@

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
	\begin{mathpar}
		{\color{blue}
			\forall \  m. \left\langle
			\begin{array}{c}
				\texttt{pn} \mapsto (p, n) \ \ast \ \nodeboxrep(n) \ \big| \ \treerep\ m
			\end{array}
			\right\rangle
		}
		\vspace{-6pt}\\ \vspace{-6pt}
		\texttt{traverse(pn, k)} 
		\\
		{\color{blue}
			\left\langle \mathit{res}. \ \exists \  n'.
			\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast \ \texttt{k} \in \mathsf{range}\ \ast\ 
				\\ 
				\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}
			\ \Bigg| \ \treerep\ m \
			\right\rangle
		}
	\end{mathpar}
As before, the \texttt{traverse} function will navigate the structure and return a node $n'$ whose range includes \texttt{k}, as well as acquiring $n'$'s lock and returning its contents. Also as before, the Boolean $\mathit{res}$ indicates whether $n'$ is empty or contains the key \lstinline{k}.

@@

Next, we define the loop invariant for the give-up template's \texttt{traverse} function:
\begin{mathpar} \mathsf{traverse\_inv}(\texttt{pn}) \triangleq \exists \ p, r, n.\ \texttt{pn} \mapsto (p, n) \ast \ \infp (r) \ast \infp (n)
\end{mathpar}
Unlike the lock-coupling case, we do not hold any locks between iterations of the loop body, and we do not maintain the fact that \lstinline{k} is in the range of the current node (we must check this later with \lstinline{inRange}). Instead, we keep a reference $\infp(r)$ to the root node, so that we can return to it if an \lstinline{inRange} check fails.

@@

Our target data structure is a binary search tree (BST) implemented in C. A lock-coupling BST was already written verified as a demonstration of VST's concurrency capabilities~\cite{bst-conc}. We refactored it into \lstinline{findNext}, \lstinline{traverse}, \lstinline{insertOp}, etc. as required by the template approach, and then re-verified it using the template; we then replaced \lstinline{traverse} with the give-up version and verified the resulting give-up BST as well. We were able to obtain verified \lstinline{insert} and \lstinline{lookup} operations using this technique; for \texttt{delete}, we found a mismatch between the template approach and BST deletion, which we discuss in Section \ref{internal_reorganize}.

@@

\begin{figure}[h]
	\centering
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)} 
			\end{array}
			\right\}
		}
		\\ 
		\texttt{findNext(pn, k)} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  \texttt{n'.}
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto \texttt{(k', v', l, r)}  \\\  
				\ast \ (\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (\texttt{l} = \texttt{n'} \land \texttt{k} < \texttt{k'}) \lor (\texttt{r} = \texttt{n'} \land \texttt{k} > \texttt{k'})\ \mathsf{else}\ (\texttt{n} = \texttt{n'} \land \texttt{k} = \texttt{k'}))
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure}

@@

The contents of a BST node are defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node {int key; void *value; struct node_t *left, *right;} node;
\end{lstlisting}
The BST's \lstinline{findNext} function (Figure \ref{findnext}) indicates whether the target key is in the current node, its left subtree, or its right subtree. The precondition of \texttt{findNext} (Figure \ref*{fig:findNext_lock}) states that the input has a non-null \texttt{t} field pointing to a \texttt{node} struct with key \texttt{k'}, value \texttt{v'}, and pointers to the left \texttt{l} and right \texttt{r} child nodes. The postcondition describes three possible outcomes: either the provided key \texttt{k} is less than the node's key \texttt{k'}, and the next node \lstinline{n'} is the left child \lstinline{l}; \lstinline{k} is greater than \lstinline{k'} and \lstinline{n'} is the right child \lstinline{r}; or \lstinline{k} is exactly \lstinline{k'} and \lstinline{n'} is the original node \lstinline{n}, in which case the return value is 0. The proof is straightforward.

\begin{figure}[ht]
	\lstinputlisting[language=C, style=myStyle]{findnext.c} 
	\caption{C implementation of \lstinline{findNext} for the binary search tree (all templates)}
	\label{findnext}
\end{figure}
